Compilation through interpretation

### 2023-01-06 Papers/ENASE-2023/c-=-1.tex
fixes

\usepackage{epsfig}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{apalike}
\usepackage[bottom]{footmisc}
\usepackage{SCITEPRESS}     % Please add other packages that you may need BEFORE the SCITEPRESS.sty package.

\begin{document}

\title{Authors' Instructions: Preparation of Camera-Ready Contributions to SCITEPRESS Proceedings}

\author{
	\authorname{
		Adam Grabski\sup{1}
		\orcidAuthor{0000-0000-0000-0000}, Ilona Bluemke\sup{1}\orcidAuthor{0000-0000-0000-0000}
	}
	\affiliation{\sup{1}Department of Electronics and Information Technology, Warsaw University of Technology, Nowowiejska 15/19, Warsaw, Poland}
	\email{adam.grabski@stud.pw.edu.pl, Ilona.Bluemke@pw.edu.pl}
}

\keywords{The paper must have at least one keyword. The text must be set to 9-point font size and without the use of bold or italic font style. For more than one keyword, please use a comma as a separator. Keywords must be titlecased.}

\abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words. The text must be set to 9-point font size.}

@@

%% Bibliography
\bibliography{bibliography}

\end{document}

>>>>

\usepackage{epsfig}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{apalike}
\usepackage[bottom]{footmisc}
\usepackage{SCITEPRESS}     % Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
\lstset{tabsize=2}
\begin{document}

\title{Compilation through interpretation}

\author{
	\authorname{
		Adam Grabski\sup{1}
		\orcidAuthor{0000-0002-2894-5976}, Ilona Bluemke\sup{1}\orcidAuthor{0000-0002-2894-5976}
	}
	\affiliation{\sup{1}Department of Electronics and Information Technology, Warsaw University of Technology, Nowowiejska 15/19, Warsaw, Poland}
	\email{adam.grabski@stud.pw.edu.pl, Ilona.Bluemke@pw.edu.pl}
}

\keywords{compiler, interpreter, metaprogramming}

\abstract{\input{sections/abstract.tex}}

@@

%% Bibliography

\bibliographystyle{apalike}
{\small
	\bibliography{bibliography}}
\end{document}

### 2023-01-06 Papers/ENASE-2023/sections/abstract.tex
fixes

As static metaprogramming is becoming more relevant, compilers must adapt to accommodate them. This requires exposing more information about the code, from the compiler to the programmer as well as more powerful compile-time function execution capabilities. The Interpreter component of a compiler therefore becomes more important.

@@

Compile Time Function Execution (CTFE) First pattern was designed for low-level, non-garbage-collected, reflection-enabled language C-=-1. This is a new programming language, which allows the programmer to execute any code at the compile time, as well as to analyze and modify the program structure. Grace to the extensibility of the designed compiler, programs written in C-=-1 can also generate marshalling bindings for other languages and support a variety of programming paradigms.

The significant flexibility and extensibility of CTFEF caused severe problems in compiler construction. The compiler appeared very complex, its parts, especially Interpreter, were very difficult to debug. Compiler operates on inflexible data structures, accessible also for a programmer. They are therefore a part of the compiled languages standard library and backwards compatibility must be maintained.

>>>>

However, compiler is a very important program not many of us has the chance to implement it. Currently, the chances of implementing a compiler are increasing because many domains languages are proposed, and theirs compilers has to be produced.

@@

Compile Time Function Execution (CTFE) First pattern was designed for low-level, non-garbage-collected, reflection-enabled language C-=-1. This is a new programming language, which allows the programmer to execute any code at the compile time, as well as to analyze and modify the program structure.
The significant flexibility and extensibility of CTFEF caused severe problems in compiler construction. The compiler appeared very complex, its parts, especially Interpreter, were very difficult to debug.

### 2023-01-06 Papers/ENASE-2023/sections/conclusions.tex
fixes

On the other hand, using CTFEF approach has some drawbacks. Implementing a compiler is more difficult.
Since a significant part of the compiler is interpreted, the initial implementation requires working with the target language.
The lack of available tools, such as integrated development environments, debuggers and libraries makes this part of the process significantly more difficult.
On the other hand also means the work on the compiler in the target language may start at the very beginning of compiler bootstrapping process\cite{puntambekar:compiler_design, novillo2007gcc}.

>>>>

On the other hand, using CTFEF approach has some drawbacks. Implementing a compiler is more difficult.
Since a significant part of the compiler is interpreted, the initial implementation requires working with the target language.
The lack of available tools, such as integrated development environments, debuggers and libraries makes this part of the process significantly more difficult.
On the other hand also means the work on the compiler in the target language may start at the very beginning of compiler bootstrapping process\cite{puntambekar:compiler_design,novillo2007gcc}.

### 2023-01-06 Papers/ENASE-2023/sections/design.tex
fixes

	\begin{lstlisting}[
	  numbers=left,
	  firstnumber=0,
	  caption={noDiscard attribute in C-=-1},
	  aboveskip=0pt,
	  label={lst:noDiscardCm1}
	  ]
public att<function> NoDiscard
{
	public fn attach(f: functionDescriptor)
	{}
	public fn onCall(call: functionCallExpression*)
	{
		if(call._parentStatment != null<IInstruction>())
			raiseError(
			&(call._pointerToSource),
			"Return value of a no-discard function is not used",
			123
			);
}
}
  \end{lstlisting}
\end{minipage}

@@

	\begin{lstlisting}[
	  numbers=left,
	  firstnumber=0,
	  caption={Example of using noDiscard attribute from Listing \ref{lst:noDiscardCm1}},
	  aboveskip=0pt,
	  label={lst:noDiscardUsageCm1}
	  ]
  [noDiscard()]
  fn noDiscardFunction() -> usize;
  fn main() -> usize {
	noDiscardFunction();
	// error 123: Return value of
	// a no-discard function is not used
	let x = noDiscardFunction();     // ok
	let y = x + noDiscardFunction(); // ok
	return noDiscardFunction();      // ok
  }
  \end{lstlisting}
\end{minipage}

@@

What is unique about CTFEF is that this part of the compiler can be written in the target language, during initial bootstrapping of the compiler bootstrapping process, i.e. initial implementation using another language \cite{puntambekar:compiler_design, novillo2007gcc}.
In case of the C-=-1 compiler, C++ was used to implement Frontend, Backend and Interpreter, with Compiler Interface written in C-=-1 \cite{grabski2022compilation}.

>>>>

	\begin{lstlisting}[
		numbers=left,
		tabsize=2,
		firstnumber=0,
		breaklines=true,
		caption={noDiscard attribute in C-=-1},
		aboveskip=0pt,
		label={lst:noDiscardCm1},basicstyle=\tiny
	]
public att<function> NoDiscard
{
	public fn attach(f: functionDescriptor)
	{}
	public fn onCall(call: functionCallExpression*)
	{
		if(call._parentStatment != null<IInstruction>())
			raiseError(
			&(call._pointerToSource),
			"Return value of a no-discard function is not used",
			123
			);
	}
}
  \end{lstlisting}
\end{minipage}

@@

	\begin{lstlisting}[
		numbers=left,
		firstnumber=0,
		caption={Example of using noDiscard attribute from Listing \ref{lst:noDiscardCm1}},
		aboveskip=0pt,
		breaklines=true,
		label={lst:noDiscardUsageCm1},basicstyle=\tiny
	  ]
  [noDiscard()]
  fn noDiscardFunction() -> usize;
  fn main() -> usize {
	noDiscardFunction();
	// error 123: Return value of
	// a no-discard function is not used
	let x = noDiscardFunction();     // ok
	let y = x + noDiscardFunction(); // ok
	return noDiscardFunction();      // ok
  }
  \end{lstlisting}
\end{minipage}

@@

What is unique about CTFEF is that this part of the compiler can be written in the target language, during initial bootstrapping of the compiler bootstrapping process, i.e. initial implementation using another language \cite{puntambekar:compiler_design,novillo2007gcc}.
In case of the C-=-1 compiler, C++ was used to implement Frontend, Backend and Interpreter, with Compiler Interface written in C-=-1 \cite{grabski2022compilation}.

### 2023-01-06 Papers/ENASE-2023/sections/implementation.tex
fixes

\begin{figure*}
	\includegraphics[width=16cm]{pictures/Interpreter_data_structures_uml.png}
	\caption{Class diagram of C-=-1 Interpreter data structures}
	\label{fig:Interpreter_data_structures}
\end{figure*}

@@

@@

@@

\begin{minipage}{\linewidth}
	\begin{lstlisting}[
	tabsize=2,
	numbers=left,
	stepnumber=1,
	caption={Representation of average function in C-=-1 and LLVM IR},
	label={impl:theoretical_llvmir}
	]
// C-=-1 function in source code
fn average(a: usize, b: usize, c: usize) -> usize {
	return (a + b + c) / 3;
}
// Ideal representation in LLVMIR
define i32 @average(i32 %0, i32 %1, i32 %2){
  %4 = add i32 %1, %0
  %5 = add i32 %4, %2
  %6 = sdiv i32 %5, 3
  ret i32 %6
}
// Generated LLVM IR
define i32 @__operator___plus____usize__usize(i32 %0, i32 %1){
	%3 = add i32 %1, %2
	ret i32 %3
}
define i32 @__operator___div____usize__usize(i32 %0, i32 %1){
	%3 = sdiv i32 %1, %2
	ret i32 %3
}
define i32 @average(i32 %0, i32 %1, i32 %2){
	%4 = call __operator___plus____usize__usize(i32 %1, i32 %0)
	%5 = call __operator___plus____usize__usize(i32 %4, i32 %2)
	%6 = call __operator___div____usize__usize (i32 %5, i32 3)
	ret i32 %6
}
\end{lstlisting}
\end{minipage}

Functions \lstinline{buildFunction} and \lstinline{getFunctionName}, from Listing \ref{lst:selected_Backend_interface}, are parts of Backend Interface.
They are invoked in order to convert a C-=-1 \lstinline{functionDescriptor} to a LLVMIR function.
They were included in the Listing, because they are the only parts of the Backend Interface that need to interact with \lstinline{ISymbolNameOverride} attributes.
Both of these procedures, check whether an attribute implementing this interface is attached to the function they are currently processing.
This happens on lines three and nineteen of Listing \ref{lst:selected_Backend_interface}.
If that attribute is present, code of that function is ignored, and it is treated as an external symbol: condition on line twenty omits execution of \lstinline{build_block} function on line twenty-three.
Procedure \lstinline{getFunctionName} contains a similar condition on line three, that decides how the name should be generated.
If an \lstinline{ISymbolNameOverride} attribute is present, it will be created by \lstinline{createSymbolName} method of the attribute attached to the function.
Otherwise, the \lstinline{mangleName} function will create name of function's symbol based on its parameters and return type.

@@

	\begin{lstlisting}[
	  numbers=left,
	  firstnumber=0,
	  caption={C-=-1 memory allocation functions from standard library},
	  aboveskip=0pt,
	  label={lst:new}
	  ]
[mapToExternalSymbol("malloc", "")]
private fn unsafe_new(size: usize) -> char* {}

	\begin{lstlisting}[
	  numbers=left,
	  firstnumber=0,
	  caption={Part of a Backend Interface, using \lstinline{ISymbolNameOverride} interface},
	  aboveskip=0pt,
	  label={lst:selected_Backend_interface}
	  ]

  private fn getFunctionName(f: functionDescriptor) -> string {
	let attribute = f.get_attribute<ISymbolNameOverride>();
	if(attribute != null<ISymbolNameOverride>())
	  return attribute.createSymbolName();
	return mangleName(f);
  }

  private fn buildFunction(
	f: functionDescriptor,
	llvmF: llvmFunction,
	registry: packageRegistry*,
	mod: llvmModule)
  {
	let variables = dictionary<variableDescriptor, llvmValue>();
	let params = f.parameters();
	for(i in enumerate(0, params.length()))
	  variables.push(params[i], llvmF.getParameter(i));
	let builder = llvmF.getBuilder();
	let attribute = f.get_attribute<ISymbolNameOverride>();
	if(attribute == null<ISymbolNameOverride>())
	{
	  let code = f.code();
	  build_block(&code, &builder, &variables, registry);
	}
  }

>>>>

\begin{figure*}
	\includegraphics[width=15cm]{pictures/Interpreter_data_structures_uml.png}
	\caption{Class diagram of C-=-1 Interpreter data structures}
	\label{fig:Interpreter_data_structures}
\end{figure*}

@@

\begin{minipage}{\linewidth}
	\begin{lstlisting}[
	tabsize=2,
	numbers=left,
	stepnumber=1,
	breaklines=true,
	caption={Representation of average function in C-=-1 and LLVM IR},
	label={impl:theoretical_llvmir},basicstyle=\tiny
	]
// C-=-1 function in source code
fn average(a: usize, b: usize, c: usize) -> usize {
	return (a + b + c) / 3;
}
// Ideal representation in LLVMIR
define i32 @average(i32 %0, i32 %1, i32 %2){
  %4 = add i32 %1, %0
  %5 = add i32 %4, %2
  %6 = sdiv i32 %5, 3
  ret i32 %6
}
// Generated LLVM IR
define i32 @__operator___plus___usize__usize(i32 %0, i32 %1){
	%3 = add i32 %1, %2
	ret i32 %3
}
define i32 @__operator___div___usize__usize(i32 %0, i32 %1){
	%3 = sdiv i32 %1, %2
	ret i32 %3
}
define i32 @average(i32 %0, i32 %1, i32 %2){
	%4 = call __operator___plus___usize__usize(i32 %1, i32 %0)
	%5 = call __operator___plus___usize__usize(i32 %4, i32 %2)
	%6 = call __operator___div___usize__usize (i32 %5, i32 3)
	ret i32 %6
}
\end{lstlisting}
\end{minipage}

@@

	\begin{lstlisting}[
		numbers=left,
		firstnumber=0,
		breaklines=true,
		caption={C-=-1 memory allocation functions from standard library},
		aboveskip=0pt,
		label={lst:new},basicstyle=\tiny
	  ]
[mapToExternalSymbol("malloc", "")]
private fn unsafe_new(size: usize) -> char* {}

@@

Functions \lstinline{buildFunction} and \lstinline{getFunctionName}, from Listing \ref{lst:selected_Backend_interface}, are parts of Backend Interface.
They are invoked in order to convert a C-=-1 \lstinline{functionDescriptor} to a LLVMIR function.
They were included in the Listing, because they are the only parts of the Backend Interface that need to interact with \lstinline{ISymbolNameOverride} attributes.
Both of these procedures, check whether an attribute implementing this interface is attached to the function they are currently processing.
This happens on lines two and eighteen of Listing \ref{lst:selected_Backend_interface}.
If that attribute is present, code of that function is ignored, and it is treated as an external symbol: condition on line eighteen omits execution of \lstinline{build_block} function on line twenty-one.
Procedure \lstinline{getFunctionName} contains a similar condition on line two, that decides how the name should be generated.
If an \lstinline{ISymbolNameOverride} attribute is present, it will be created by \lstinline{createSymbolName} method of the attribute attached to the function.
Otherwise, the \lstinline{mangleName} function will create name of function's symbol based on its parameters and return type.

@@

	\begin{lstlisting}[
		numbers=left,
		breaklines=true,
		firstnumber=0,
		caption={Part of a Backend Interface, using \lstinline{ISymbolNameOverride} interface},
		aboveskip=0pt,
		label={lst:selected_Backend_interface},basicstyle=\tiny
	  ]
private fn getFunctionName(f: functionDescriptor) -> string {
let attribute = f.get_attribute<ISymbolNameOverride>();
if(attribute != null<ISymbolNameOverride>())
	return attribute.createSymbolName();
return mangleName(f);
}
private fn buildFunction(
f: functionDescriptor,
llvmF: llvmFunction,
registry: packageRegistry*,
mod: llvmModule)
{
	let variables = dictionary<variableDescriptor, llvmValue>();
	let params = f.parameters();
	for(i in enumerate(0, params.length()))
		variables.push(params[i], llvmF.getParameter(i));
	let builder = llvmF.getBuilder();
	let attribute = f.get_attribute<ISymbolNameOverride>();
	if(attribute == null<ISymbolNameOverride>())
	{
		let code = f.code();
		build_block(&code, &builder, &variables, registry);
	}
}
  \end{lstlisting}
\end{minipage}

### 2023-01-06 Papers/ENASE-2023/sections/introduction.tex
fixes

Compile-time function execution (CTFE) is an aspect of a programming language, that allows the programmer to execute code at compile-time.
It has been gaining popularity with low level programming languages such as Rust \cite{rust} or C++ \cite{ISO:cpp98}.
The goals and capabilities of CTFE depend on the language and are discussed in details in chapter \ref{related-work}.

>>>>

Software Engineering is concerned with building good software. Very important type of software are compilers. Many books on compilers constructions has been published, among them the books written  by team Aho, Sethi, Ullman eg. \cite{aho2007compilers} are very popular. Currently, the chances of implementing a compiler are increasing because many domains languages are proposed and theirs compilers has to be produced. As new languages have new features theirs compilers must differ from the classic ones.

Compile-time function execution (CTFE) is an aspect of a programming language, that allows the programmer to execute code at compile-time.
It has been gaining popularity with programming languages without virtual machines such as Rust \cite{rust} or C++ \cite{ISO:cpp98}.
The goals and capabilities of CTFE depend on the language and are discussed in details in chapter \ref{related-work}.
Introduction of such mechanism to a language, significantly affects how software is written.

### 2023-01-06 Papers/ENASE-2023/sections/realted-work.tex
fixes

Many currently used programming languages allow the programmer to execute some code at compile time, for example: C\# \cite{csharp:source_generators,roslyn}, Rust \cite{rust, klabnik2019rust} and C++ \cite{ISO:cpp20}.

>>>>

Many currently used programming languages allow the programmer to execute some code at compile time, for example: C\# \cite{slimaksource,roslyn}, Rust \cite{rust,klabnik2019rust} and C++ \cite{ISO:cpp20}.
